---
title: "Activities - Scatterplot, Correlation, Crime"
author: "Wells"
date: "2025-10-07"
output: html_document
Dariana Guzman
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Activity #1:

Examine this New York Times scatterplot from 2013:
https://archive.nytimes.com/www.nytimes.com/interactive/2013/05/07/education/college-admissions-gap.html?_r=0

The graphic displays four different scatterplots showing relationships between student demographics and college acceptance rates. For each of the four scatterplots:

a) Describe the relationship (or lack thereof) between the two variables.
b) Note the specific words and phrases the graphic uses to characterize each relationship.

Write 1-2 sentences summarizing each scatterplot.

Name each person in your group who worked on this entry.
Eden, Owen, Mara + others

**Write Answer #1:**

Scatterplot 1
The relationship of the scatterplot entails that around 15% of black students that attend elite/competitive schools. There seems to be a positive correlation, because out of the amount of students that do attend these universities -- the majority do graduate. 

Scatterplot 2
When looking at the bigger picture, you can say that there is a negative correlation. Because when this happens, you can see that many more black students attend colleges with low graduation rates. One example is Baltimore City Community College where 88.7% of black students attend a 3.2% graduation rate. 

Scatterplot 3
Negative correlation. Hispanic students are more likely to attend community colleges, which can lead to more dropouts. They are the most underrepresented at most elite colleges. An example can be Baton Rouge Community College that has 100% Hispanic students with a 4.6% graduation rate. They have a low graducation rate compared to the percentage of students. 

Scatterplot 4
Positive correlation.Asian students outperform. More Asian students go to more elite/competitive schools, so they have a high graduation rate with a larger amount of students. 

-------------------------------------------------------------------------------

Activity #2: Correlation Demonstration

In the following exercise, we will look at data about 16 different consultants, their aptitude scores and their performance scores.

Our question: what is the relationship between a consultant's aptitude and their performance? 

How are they correlated?

Below we calculate a Pearson's correlation coefficient and then create a regression line.

```{r}
library(tidyverse)
```

Load data for aptitude and performance for 16 different consultants
```{r}
aptitude <- c(45, 81, 65, 87, 68, 91, 77, 61, 55, 66, 82, 93, 76, 83, 61, 74)
performance <- c(56, 74, 56, 81, 75, 84, 68, 52, 57, 82, 73, 90, 67, 79, 70, 66)
consultant <- c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16)
#this builds our table
consultants <- data.frame(consultant, aptitude, performance)
consultants
```


### Task 1: Use `summarize()` and the `cor()` function to calculate the correlation between aptitude and performance, using the pearson method.

```{r}
consultants |> summarize(correlation = cor(aptitude, performance, method="pearson"))
```

We see a Pearson's correlation coefficient of 0.783.
What does that mean?

The Pearson correlation scale is from -1 to 1.
-1 is a perfect negative correlation
1 is a perfect positive correlation

How do we interpret the values in between?

-- High Degree: Values between ±0.50 and ±1 suggest a strong correlation.
-- Moderate Degree: Values between ±0.30 and ±0.49 indicate a moderate correlation.
-- Low Degree: Values below +0.29 are considered a weak correlation.
-- No Correlation: A value of zero implies no relationship.

**Question: How would you describe the 0.783 correlation between aptitute and performance?**

**A: Write your answer:**

A correlation of 0.783 between aptitude and performance indicates a strong positive relationship. This can mean that as aptitude increases, performance will most likely increase as well.

### Task 2: Create a scatterplot of aptitude and performance, using `geom_point()`

```{r}

consultants |> 
  ggplot() +
  geom_point(aes(x=aptitude, y=performance)) +
  geom_smooth(aes(x=aptitude, y=performance), method="lm") +
  expand_limits(x = 0, y = 0)

```


The "best fit" line is known as a regression line that helps us understand the relationship between two variables. This predicts the dependent variable based on the independent variable. 

A regression line is a straight line showing a best-fit between independent and dependent variables.

In the example above, aptitude would be the independent variable, or the predictor. That's because we would expect aptitude to predict performance, which is the dependent variable, or the outcome variable.

Independent = predictor
Dependent = outcome

Regression is one step ahead of correlation. It helps predict the values of dependent variables or outcomes with respect to changes in independent or explanatory variables. 

Regression lines help predict outcomes.


-------------------------------------------------------------------------------

Activity #3: 

# PG Murders and Heat

Let's explore the relationship between crime and hot weather using homicide and temperature data from Prince George's County. 

You'll load the data, perform some basic statistical analysis, and create a scatterplot to explore the relationship. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

**write code below**
```{r}
library(tidyverse)
```


### Task 2: Load the dataset from [https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_murder_temps.csv](https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_murder_temps.csv)

**write code below**
```{r}

pg_murder_temps <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_murder_temps.csv")

```



### Task 3: Use `summarize()` and the `cor()` function to calculate the correlation between murders and temperature, using the pearson method.

**write code below**
```{r}
pg_murder_temps <- pg_murder_temps

pg_murder_temps |> summarize(correlation = cor(murders, temperature, method="pearson"))

```



### Task 4: Create a scatterplot of murders and temperature, using `geom_point()`. Your aes() will have temperatures on the x axis, murders on the y axis and add a label for the month (so label=month inside the aes()).

**write code below**
```{r}
pg_murder_temps |> 
  ggplot() +
  geom_point(aes(x=temperature, y=murders, label = month)) 




```


### Task 5: Add labels to the scatterplot using geom_text(). Inside geom_text() put the same aes() as geom_point in Task 4, then after the aes() add size = 5.

**write code below**
```{r}
pg_murder_temps |> 
  ggplot() +
  geom_point(aes(x = temperature, y = murders)) +
  geom_text(aes(x = temperature, y = murders, label = month), size = 5)

```


### Task 6: Describe the results of both the correlation calculation and the scatterplot.

**Write Answer:**

There seems to be a moderate positive correlation between the amount of murders and the temperature. The calculated correlation was 0.58, which is not necessarily that close to 1 but also not close to 0. When we take a look at the scatterplot, you can see that there is that moderate positive correlation. During warmer months, where the temp is a bit higher -- there seems to be more murders recorded than when it is cold. But it all just depends because we can see that August was semi low compared to the months of Jul, Jun, , and May. You can see where the uncertainty is and why the calculation turned out the way it did. So you can't just claim that one season will be worse in terms of murder if there isn't consistency. 

(I feel like I explained horribly but it does make sense in my head, I can see why everything turned out the way it did)

When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
