---
title: "Crime Rates"
Name: Dariana Guzman
---

Follow along with the demo to learn how to calculate crime rates using R and the Tidyverse. We'll also introduce a new library, janitor, which helps make column names easier to work with. You should get started by loading the tidyverse in a new R block, then install and load janitor.

### Task 1: Load the tidyverse and install and load janitor

```{r}
library(tidyverse)
install.packages("janitor")
library(janitor)
```



### Task 2: Load the crime dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_annual_crime.csv

```{r}
crimes <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_annual_crime.csv") |> clean_names()
```


### Task 3: Use janitor to clean up the column names 

ROB WELLS SAID TO IGNORE - ALREADY DONE

### Task 4: Calculate the total crime rate for each year (picking our unit!)

```{r}
crimes |> mutate(total_rate = grand_total/population * 10000)
```



### Task 5: Calculate the violent and property crime rates for each year (same unit)

```{r}
crimes_with_rates <- crimes |> mutate(violent_rate = violent_crime_total/population * 10000, property_rate = property_crime_total/population * 10000)
```



### Task 6: Describe our results

Write a few sentences about what you found in the data. What years had the highest crime rates? What years had the lowest crime rates? Does the pattern match your expectations?

The year with the highest violent crime rate is 1992 with a rate of 99.91850.
The year with the highest property crime rate is 1980 with a property crime rate of 577.5091

The year with the lowest violent crime rate is 2020 with a rate of 41.22480
The year with the lowest property crime rate is 2020 with a property crime rate of 162.6187

This perfectly matches my expectations considering that 2020 was a year where the COVID-19 pandemic spread across the globe, causing everyone to go on lock down. In order to avoid contamination, most people stayed in -- so the majority of people did. With the exceptions of when trying to buy necessities and in this case, those committing crimes. It makes sense why the rates look the way they do. 

I also noticed that the trend of crime was already going down in the 2010s before COVID hit. 



When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.

