---
title: "Diversity Index for Maryland Cities"
date: "Nov 8, 2025"
author: "Your Name"
---

**Your Name Here**
Dariana Guzman

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Overview

In this exercise, we will calculate the **USA TODAY Diversity Index** for Maryland cities with a population of at least 10,000 using 2020 Census race and ethnicity data. The Diversity Index reflects the probability that two randomly selected individuals from a city are of different races or ethnicities.

A score of 0 indicates no diversity (everyone is the same race/ethnicity), while a score closer to 100 indicates higher diversity. The **most diverse city** has the highest index, reflecting a balanced mix of several racial and ethnic groups. The **least diverse** city likely has one overwhelmingly dominant racial/ethnic group.

## Task 1: Load and Describe the Data (3 points)

Run the following code to load the data, then examine it and describe what it contains. Based on the raw numbers alone, which cities stand out as diverse to you?



```{r load-data}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/usat_diversity_index.csv")

# Preview structure
View(md_cities)
```

**ANSWER HERE**

The dataset lists 128 towns and cities in Maryland and shows how many people in each place belong to different racial and ethnic groups. It includes totals for groups like White, Black, Asian, Native American, and Hispanic residents. Some places, such as Silver Spring, Gaithersburg, Germantown, Columbia, and Wheaton, stand out as very diverse because no single group makes up most of the population. Others, like Bowie and Clinton, are mostly Black, while towns such as Severna Park and Bel Air are mostly White. Areas like Langley Park and Chillum have large Hispanic populations, showing how different parts of Maryland have very different communities.


## Task 2: Calculate Race Percentages

```{r race-percentages}
md_cities <- md_cities |>
  mutate(
    White_pct = White / TotalPop,
    Black_pct = Black / TotalPop,
    Native_pct = Native / TotalPop,
    Asian_pct = Asian / TotalPop,
    NHPI_pct = NHPI / TotalPop
  )
```

## Task 3: Square the Race Percentages

We need to square the race percentages in order to calculate the probability that any two people chosen at random from a given census area are of different races or are Hispanic or not.

```{r square-race}
md_cities <- md_cities |>
  mutate(
    White_pct_sq = White_pct^2,
    Black_pct_sq = Black_pct^2,
    Native_pct_sq = Native_pct^2,
    Asian_pct_sq = Asian_pct^2,
    NHPI_pct_sq = NHPI_pct^2
  )
```


## Task 4: Calculate Race Probabilities (3 points)

The race probability is the sum the squared probabilities for each of the races. It represents the probability that two people randomly chosen from the same city will be of the same race. Which city has the highest race probability, and why do you think that is?


```{r race-prob}
md_cities <- md_cities |>
  mutate(
    Race_Prob = White_pct_sq + Black_pct_sq + Native_pct_sq + Asian_pct_sq + NHPI_pct_sq
  ) |> 
  arrange(desc(Race_Prob))

View(md_cities)
```

**ANSWER HERE**

Cities like Fort Washington, Kemp Mill, and Seabrook stand out as some of the most diverse. No single racial group dominates the population, so people from different racial backgrounds are more evenly represented. In contrast, places like Summerfield or Lake Shore are less diverse because one group -- either white or Black residents -- makes up most of the community.

But overall, the city with the highest race probability is Summerfield, since it’s the least racially diverse. This happens because one racial group makes up almost the entire population, most residents are Black here. 

## Task 5: Calculate Ethnicity Percentages and Their Squares (3 points)

Hispanic origin is not a race but an ethnicity or nationality – and a separate Census question. The probability someone is Hispanic or not Hispanic must be calculated separately from race, and Hisp_Prob is the probability that two randomly selected people will be Hispanic or not. That's an important distinction, because unlike race probability, this measures the likelihood of a binary question: that both random people will be Hispanic. What do you notice about the cities with the *lowest* Hisp_Prob scores?


```{r ethnicity}
md_cities <- md_cities |>
  mutate(
    HispTotal = HISPANIC + NotHispanic,
    Hisp_pct = HISPANIC / HispTotal,
    NotHisp_pct = NotHispanic / HispTotal,
    Hisp_pct_sq = Hisp_pct^2,
    NotHisp_pct_sq = NotHisp_pct^2,
    Hisp_Prob = Hisp_pct_sq + NotHisp_pct_sq
  ) |> 
  arrange(Hisp_Prob)

View(md_cities)
```

**ANSWER HERE**

The cities with the lowest Hisp_Prob scores, such as Langley Park, Chillum, and Hyattsville, have a nearly even mix of Hispanic and non Hispanic residents. This balance lowers the probability that two randomly chosen people will share the same Hispanic background. These areas are usually more ethnically diverse, with strong representation from both Hispanic and non-Hispanic communities.


## Task 6: Multiply Race and Ethnicity Probabilities

Here we multiply the two probabilities together to get the probability that any two random people have the SAME race and ethnicity.

```{r combined-prob}
md_cities <- md_cities |>
  mutate(
    RaceEth_Prob = Race_Prob * Hisp_Prob
  )
```


## Task 7: Calculate Diversity Index (3 points)

The final step is to calculate the probability that two people randomly chosen WILL NOT BE OF THE SAME RACE OR ETHNICITY. In other words, the highest index scores will be cities where those two random people will be from different backgrounds. Put simply, more diverse. Which cities are the top 3, and where are they located? Look at the bottom of the list - which city is the least diverse, and where is it?


```{r diversity-index}
md_cities <- md_cities |>
  mutate(
    Diversity_Index = (1 - RaceEth_Prob) * 100
  ) |> 
  arrange(desc(Diversity_Index))

View(md_cities)
```

**ANSWER HERE**

Most diverse:

Langley – DI: 97.18 – PG County

Wheaton – DI: 90.88 – Montgomery County

Adelphi – DI: 90.04 – PG County

Least diverse:

Lake Shore – DI: 15.53 – Anne Arundel County

### Pitch a story (8 points)

Given the data that you've produced here, write a headline and lede of the most interesting story from this exercise, and add below that details on how you would report it out: who you'd talk to and the questions you'd ask. When you describe the Diversity Index values, do so in a similar way to how USA Today did; the raw values themselves won't mean much to readers.

**ANSWER HERE**

Headline: Langley Park ranks as Maryland’s most diverse city, diversity index shows

Lede: Langley Park has the state’s most diverse population, with residents from many racial and ethnic backgrounds populating the area. By contrast, Lake Shore has the least diverse population, according to data.

To report this story, I would talk to local residents in both Langley Park and Lake Shore to hear what it’s like living in these communities. 

I would also ask community leaders about programs that bring people of different backgrounds together and talk to experts about why some cities are more diverse than others. 

Questions could include how diversity affects schools, neighborhoods, and daily life, and whether city policies have influenced who lives there. 

By describing the cities this way, readers can understand which communities are most mixed and which aren't as much without needing to know the raw numbers.

