---
title: "Final Project"
output: html_document
DARIANA GUZMAN
---

For this project, I am using the College Scorecard dataset. It is a large dataset from the U.S. Department of Education. It includes information about colleges such as tuition, test scores, graduation rates, and student earnings. I am focusing on tuition costs at colleges in Maryland and Virginia.

---
#Where the data comes from?#

The data comes from the College Scorecard, which is public information from the federal government.

This project uses the US College Scorecard dataset, accessed through the `collegeScorecard` R package.

Main website: https://collegescorecard.ed.gov/data
R package used: collegeScorecard
Package info: https://raw.githubusercontent.com/wellsdata/Jour405_student_repo_fall_2025/refs/heads/main/project/college_scorecard_2021_2024.csv

The package gives me two datasets:

school – information about each college (name, state, type of school, etc)
scorecard – information for each college for every year (tuition, average costs, test scores, etc)

---
```{r}
library(tidyverse)

college_data <- read_csv("https://raw.githubusercontent.com/wellsdata/Jour405_student_repo_fall_2025/refs/heads/main/project/college_scorecard_2021_2024.csv")

glimpse(college_data)
```



#What does the data contain?#

The dataset includes things like:

in-state tuition, out-of-state tuition, in and out of state enrollment

The data covers 3 school years from 2021-2024, because each school appears once per academic year.

#Most important?#

Since I am studying college tuition in Maryland and Virginia, the most important columns are:
	
institution - the school’s name
state - needed to filter for MD and VA
xxxxtuitionfee_in - year in-state tuition
xxxxtuitionfee_out -year  out-of-state tuition
xxxxundergrad_enrollment - year undergrad enrollment amount

#Whats missing?#

The dataset does not include inflation adjustments and some years have missing tuition values. The data also does not include details like major/program tuition and it doesn’t explain why tuition goes up or down.

```{r}
college_data %>%
  summarize(
    mean_in = mean(`2021_22_tuitionfee_in`, na.rm = TRUE),
    median_in = median(`2021_22_tuitionfee_in`, na.rm = TRUE),
    min_in = min(`2021_22_tuitionfee_in`, na.rm = TRUE),
    max_in = max(`2021_22_tuitionfee_in`, na.rm = TRUE)
  )
```	

#Hypotheses#

1. Maryland colleges may cost less on average than Virginia colleges.
2. Public colleges in both states might be raising tuition faster than private colleges.
3. The gap between in-state and out-of-state tuition might be growing over time.

#Questions hopefully i can answer throughout the project?#

Are tuition increases steady, or do they spike in certain years?

Do Maryland and Virginia follow the same trend, or are they different?

Do private schools raise tuition more often? Compare that to public schools?

Should I adjust tuition for inflation?

#Other Questions?#

How do I know if something I find is actually “newsworthy”?

How can I tell if my hypothesis is realistic or if I need to change it?

Should I use additional data?


