---
title: "Final Project"
output: html_document
DARIANA GUZMAN
---

For this project, I am using the College Scorecard dataset. It is a large dataset from the U.S. Department of Education. It includes information about colleges such as tuition, test scores, graduation rates, and student earnings. I am focusing on tuition costs at colleges in Maryland and Virginia.

---
#Where the data comes from?#

The data comes from the College Scorecard, which is public information from the federal government.

This project uses the US College Scorecard dataset, accessed through the `collegeScorecard` R package.

Main website: https://collegescorecard.ed.gov/data
R package used: collegeScorecard
Package info: https://raw.githubusercontent.com/wellsdata/Jour405_student_repo_fall_2025/refs/heads/main/project/college_scorecard_2021_2024.csv

The package gives me two datasets:

school – information about each college (name, state, type of school, etc)
scorecard – information for each college for every year (tuition, average costs, test scores, etc)

---
```{r}
library(tidyverse)

college_data <- read_csv("https://raw.githubusercontent.com/wellsdata/Jour405_student_repo_fall_2025/refs/heads/main/project/college_scorecard_2021_2024.csv")

glimpse(college_data)
```



#What does the data contain?#

The dataset includes things like:

in-state tuition, out-of-state tuition, in and out of state enrollment

The data covers 3 school years from 2021-2024, because each school appears once per academic year.

#Most important?#

Since I am studying college tuition in Maryland and Virginia, the most important columns are:
	
institution - the school’s name
state - needed to filter for MD and VA
xxxxtuitionfee_in - year in-state tuition
xxxxtuitionfee_out -year  out-of-state tuition
xxxxundergrad_enrollment - year undergrad enrollment amount

#Whats missing?#

The dataset does not include inflation adjustments and some years have missing tuition values. The data also does not include details like major/program tuition and it doesn’t explain why tuition goes up or down.

```{r}
college_data %>%
  summarize(
    mean_in = mean(`2021_22_tuitionfee_in`, na.rm = TRUE),
    median_in = median(`2021_22_tuitionfee_in`, na.rm = TRUE),
    min_in = min(`2021_22_tuitionfee_in`, na.rm = TRUE),
    max_in = max(`2021_22_tuitionfee_in`, na.rm = TRUE)
  )
```	

#Hypotheses#

Hypothesis 1 - MD vs VA tuition - In state:

Null: The mean in-state tuition at Maryland colleges is equal to the mean in-state tuition at Virginia colleges.

Alt: The mean in-state tuition at Maryland colleges is different from the mean in-state tuition at Virginia colleges.

Hypothesis 2 - MD vs VA tuition - Out of State:

Null: The mean out-of-state tuition at Maryland colleges is equal to the mean out-of-state tuition at Virginia colleges.

Alt: The mean out-of-state tuition at Maryland colleges is different from the mean out-of-state tuition at Virginia colleges.

#Questions hopefully i can answer throughout the project?#

Are tuition increases steady, or do they spike in certain years?

Do Maryland and Virginia follow the same trend, or are they different?

Should I adjust tuition for inflation?

#Other Questions?#

How do I know if something I find is actually “newsworthy”?

How can I tell if my hypothesis is realistic or if I need to change it?

Should I use additional data?


```{r}
md_va_colleges <- college_data %>%
  filter(state %in% c("MD", "VA")) %>%
  select(institution, state, 
         `2021_22_tuitionfee_in`, `2022_23_tuitionfee_in`, `2023_24_tuitionfee_in`) %>%
  drop_na(`2021_22_tuitionfee_in`, `2022_23_tuitionfee_in`, `2023_24_tuitionfee_in`) %>%
  mutate(
    tuition_change = `2022_23_tuitionfee_in` - `2021_22_tuitionfee_in`
  )

md_va_colleges_out <- college_data %>%
  filter(state %in% c("MD", "VA")) %>%
  select(institution, state, 
         `2021_22_tuitionfee_out`, `2022_23_tuitionfee_out`, `2023_24_tuitionfee_out`) %>%
  drop_na(`2021_22_tuitionfee_out`, `2022_23_tuitionfee_out`, `2023_24_tuitionfee_out`) %>%
  mutate(
    tuition_change_out = `2022_23_tuitionfee_out` - `2021_22_tuitionfee_out`
  )
```

T-test - In state
```{r}
tuition_md_21 <- md_va_colleges %>% filter(state == "MD") %>% pull(`2021_22_tuitionfee_in`)
tuition_va_21 <- md_va_colleges %>% filter(state == "VA") %>% pull(`2021_22_tuitionfee_in`)
t_test_21 <- t.test(tuition_md_21, tuition_va_21)

tuition_md_22 <- md_va_colleges %>% filter(state == "MD") %>% pull(`2022_23_tuitionfee_in`)
tuition_va_22 <- md_va_colleges %>% filter(state == "VA") %>% pull(`2022_23_tuitionfee_in`)
t_test_22 <- t.test(tuition_md_22, tuition_va_22)

tuition_md_23 <- md_va_colleges %>% filter(state == "MD") %>% pull(`2023_24_tuitionfee_in`)
tuition_va_23 <- md_va_colleges %>% filter(state == "VA") %>% pull(`2023_24_tuitionfee_in`)
t_test_23 <- t.test(tuition_md_23, tuition_va_23)
```	

All p-values are much greater than 0.05, which means we fail to reject the null hypothesis for all three years.

The average in-state tuition in Maryland and Virginia colleges is about the same for 2021–22, 2022–23, and 2023–24.

There is no statistically significant difference in tuition between the two states in any year.


T Test - out of state
```{r}
tuition_md_21_out <- md_va_colleges_out %>% filter(state == "MD") %>% pull(`2021_22_tuitionfee_out`)
tuition_va_21_out <- md_va_colleges_out %>% filter(state == "VA") %>% pull(`2021_22_tuitionfee_out`)
t_test_21_out <- t.test(tuition_md_21_out, tuition_va_21_out)
t_test_21_out$p.value

tuition_md_22_out <- md_va_colleges_out %>% filter(state == "MD") %>% pull(`2022_23_tuitionfee_out`)
tuition_va_22_out <- md_va_colleges_out %>% filter(state == "VA") %>% pull(`2022_23_tuitionfee_out`)
t_test_22_out <- t.test(tuition_md_22_out, tuition_va_22_out)
t_test_22_out$p.value

tuition_md_23_out <- md_va_colleges_out %>% filter(state == "MD") %>% pull(`2023_24_tuitionfee_out`)
tuition_va_23_out <- md_va_colleges_out %>% filter(state == "VA") %>% pull(`2023_24_tuitionfee_out`)
t_test_23_out <- t.test(tuition_md_23_out, tuition_va_23_out)
```

All p-values are much greater than 0.05, which means we fail to reject the null hypothesis for all three years.

The average out-of-state tuition in Maryland and Virginia colleges is about the same for 2021–22, 2022–23, and 2023–24.

There is no statistically significant difference in out of state tuition between the two states in any year.



ANOVA - MD + VA IN
```{r}
tuition_2021 <- md_va_colleges$`2021_22_tuitionfee_in`
tuition_2022 <- md_va_colleges$`2022_23_tuitionfee_in`
tuition_2023 <- md_va_colleges$`2023_24_tuitionfee_in`

anova_df <- data.frame(
  tuition = c(tuition_2021, tuition_2022, tuition_2023),
  year = factor(rep(c("2021_22", "2022_23", "2023_24"), each = nrow(md_va_colleges)))
)

anova_result <- aov(tuition ~ year, data = anova_df)
summary(anova_result)
```

ANOVA - VA IN
```{r}
va_colleges <- md_va_colleges %>% filter(state == "VA")

tuition_2021 <- va_colleges$`2021_22_tuitionfee_in`
tuition_2022 <- va_colleges$`2022_23_tuitionfee_in`
tuition_2023 <- va_colleges$`2023_24_tuitionfee_in`

anova_df <- data.frame(
  tuition = c(tuition_2021, tuition_2022, tuition_2023),
  year = factor(rep(c("2021_22", "2022_23", "2023_24"), each = nrow(va_colleges)))
)

anova_result <- aov(tuition ~ year, data = anova_df)
summary(anova_result)
```

ANOVA - MD IN
```{r}
md_colleges <- md_va_colleges %>% filter(state == "MD")

tuition_2021 <- md_colleges$`2021_22_tuitionfee_in`
tuition_2022 <- md_colleges$`2022_23_tuitionfee_in`
tuition_2023 <- md_colleges$`2023_24_tuitionfee_in`

anova_df <- data.frame(
  tuition = c(tuition_2021, tuition_2022, tuition_2023),
  year = factor(rep(c("2021_22", "2022_23", "2023_24"), each = nrow(md_colleges)))
)

anova_result <- aov(tuition ~ year, data = anova_df)
summary(anova_result)
```

The ANOVA tests whether average in state tuition changed significantly across the three years from 2021–22 to 2023–24. For Maryland, Virginia, and both states combined, all p-values are much greater than 0.05, which means we fail to reject the null hypothesis. There is no statistically significant change in in state tuition over these years, and tuition has remained relatively stable for both states.


ANOVA - MD OUT
```{r}
md_colleges_out <- md_va_colleges_out %>% filter(state == "MD")

tuition_2021_out <- md_colleges_out$`2021_22_tuitionfee_out`
tuition_2022_out <- md_colleges_out$`2022_23_tuitionfee_out`
tuition_2023_out <- md_colleges_out$`2023_24_tuitionfee_out`

anova_df_out <- data.frame(
  tuition = c(tuition_2021_out, tuition_2022_out, tuition_2023_out),
  year = factor(rep(c("2021_22", "2022_23", "2023_24"), each = nrow(md_colleges_out)))
)

anova_result_out <- aov(tuition ~ year, data = anova_df_out)
summary(anova_result_out)
```


ANOVA - VA OUT
```{r}
va_colleges_out <- md_va_colleges_out %>% filter(state == "VA")

tuition_2021_out <- va_colleges_out$`2021_22_tuitionfee_out`
tuition_2022_out <- va_colleges_out$`2022_23_tuitionfee_out`
tuition_2023_out <- va_colleges_out$`2023_24_tuitionfee_out`

anova_df_out <- data.frame(
  tuition = c(tuition_2021_out, tuition_2022_out, tuition_2023_out),
  year = factor(rep(c("2021_22", "2022_23", "2023_24"), each = nrow(va_colleges_out)))
)

anova_result_out <- aov(tuition ~ year, data = anova_df_out)
summary(anova_result_out)
```


ANOVA - MD + VA OUT
```{r}
tuition_2021_out <- md_va_colleges_out$`2021_22_tuitionfee_out`
tuition_2022_out <- md_va_colleges_out$`2022_23_tuitionfee_out`
tuition_2023_out <- md_va_colleges_out$`2023_24_tuitionfee_out`

anova_df_out <- data.frame(
  tuition = c(tuition_2021_out, tuition_2022_out, tuition_2023_out),
  year = factor(rep(c("2021_22", "2022_23", "2023_24"), each = nrow(md_va_colleges_out)))
)

anova_result_out <- aov(tuition ~ year, data = anova_df_out)
summary(anova_result_out)
```

The ANOVA tests whether average out of state tuition changed significantly across 2021–22, 2022–23, and 2023–24. For Maryland, Virginia, and both states combined, all p-values are much greater than 0.05, meaning we fail to reject the null hypothesis. Out of state tuition has remained relatively stable over these three years, with no statistically significant changes for either state or both states together.


```{r}
library(dplyr)
library(ggplot2)
library(tidyr)

bar_df <- md_va_colleges_out %>%
  select(state, `2021_22_tuitionfee_out`, `2022_23_tuitionfee_out`, `2023_24_tuitionfee_out`) %>%
  pivot_longer(
    cols = starts_with("202"),
    names_to = "year",
    values_to = "tuition"
  ) %>%
  group_by(state, year) %>%
  summarize(avg_tuition = mean(tuition, na.rm = TRUE), .groups = "drop")

ggplot(bar_df, aes(x = year, y = avg_tuition, fill = state)) +
  geom_col(position = "dodge") +
  labs(
    title = "Average Out-of-State Tuition by Year",
    x = "Academic Year",
    y = "Average Tuition ($)",
    fill = "State"
  ) +
  theme_minimal()
```
```{r}
library(dplyr)
library(ggplot2)
library(tidyr)

bar_df_in <- md_va_colleges %>%
  select(state, `2021_22_tuitionfee_in`, `2022_23_tuitionfee_in`, `2023_24_tuitionfee_in`) %>%
  pivot_longer(
    cols = starts_with("202"),
    names_to = "year",
    values_to = "tuition"
  ) %>%
  group_by(state, year) %>%
  summarize(avg_tuition = mean(tuition, na.rm = TRUE), .groups = "drop")

ggplot(bar_df_in, aes(x = year, y = avg_tuition, fill = state)) +
  geom_col(position = "dodge") +
  labs(
    title = "Average In-State Tuition by Year",
    x = "Academic Year",
    y = "Average Tuition ($)",
    fill = "State"
  ) +
  theme_minimal()
```
