---
title: "HW5_Percentiles"
Name: Dariana Guzman
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities %>% clean_names()
 
md_cities_rates <- md_cities %>%
  mutate(
    violent_rate_2019 = (violent2019 / pop2019) * 100000,
    property_rate_2019 = (property2019 / pop2019) * 100000,
    violent_rate_2020 = (violent2020 / pop2020) * 100000,
    property_rate_2020 = (property2020 / pop2020) * 100000
  )
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

ANSWER HERE
After looking at the dataframe -- Baltimore City stands out with the highest violent and property crime rates in both 2019 and 2020. This does includes the fact that Baltimore City is the most populated city though. So it's a bit obvious that if there's more people, there will be more crime. I will point out that it's rates decreased slightly in 2020. 

Other cities like Cambridge and Ocean city show high property crime rates for their population size but overall lower violent crime rates -- likely influenced by tourism or low populations that make per-capita rates appear extreme. 

The data reveals that violent crime in Maryland is mostly concentrated in urban areas, while property crime can occuer within smaller towns -- overall highlighting that crime in Maryland is unevenly distributed and influences by both city size and pop. and local factors.

### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates %>%
  mutate(
    violent_rate_change = ((violent_rate_2020 - violent_rate_2019) / violent_rate_2019) * 100,
    property_rate_change = ((property_rate_2020 - property_rate_2019) / property_rate_2019) * 100
  )
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

ANSWER HERE

Looking at the data, most cities saw violent and property crime go down from 2019 to 2020, like Baltimore, Aberdeen, and Belair. But some smaller towns, like Boonsboro, Crisfield, and Colmar Manor, had big jumps in violent crime, even if the actual number of crimes was small. The interesting part are that the outliers that tiny towns sometimes have crazy percentage changes, especially if they went from zero crimes to just one or two. It shows that while crime overall might be dropping, some small communities still see sudden spikes that stand out and make for an interesting story.


### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

ANSWER HERE

Lede: 

Maryland reports that between 2019 and 2020, there have been drops in both violent and property crime. However, several small towns have experiences spikes that stand out from the state's general trend. 

Nutgraf:

While major cities like Baltimore and Annapolis reported declines in crime rates, smaller towns such as Colmar Manor, Boonsboro, and Crisfield saw violent crime surge. Sometimes they have more than doubled even though the total number of incidents remained relatively low. These outliers show that statewide trends can hide dramatic shifts at the local level, offering a deeper look into how crime affects communities differently. Investigating these surges could reveal factors like changes in policing, local economy, or community programs, giving residents and policy makers a clearer picture of what drives crime in smaller towns.



### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

### Task 10: Take the violent crime rates for 2020 and do three things 

Using the 2020 violent crime rate, calculate the percentiles using `ntile()`. Save the results to a new dataframe called `md_cities_percentiles`. How many cities are above the 90th percentile?
```{r}
library(dplyr)

md_cities_percentiles <- md_cities_change %>%
  mutate(percentile = ntile(violent2020, 100))

md_cities_percentiles %>%
  filter(percentile > 90) %>%
  nrow()
```
Calculate the Interquartile Range of the 2020 violent crime rate using `summarize()` and R's `IQR()` function. The syntax is similar to how `mean()` works.
```{r}
md_cities_change %>%
  summarise(iqr_violent2020 = IQR(violent2020, na.rm = TRUE))
```

Do the results change your best story idea from the previous assignment? Why or why not?

No, not really. The IQR being 17 shows that most MD cities have similar, relatively low violent crime rates while there's only a handful that stand out with much higher rates. It supports my previously thought because the main takeaway is not about small shifts across most towns -- but how a few places drive up the state's image of crime (see Baltimore.) It reinforces the idea that crime is not evenly spread but significant in a few areas or hotspots. 
