---
title: "HW9_TestScores"
Name: Derek Willis/Rob Wells
Date: 10/21/2025
---

name: Dariana Guzman


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Did a New Reading Program Lead to Better Scores?

The superintendent recently claimed that a new reading program has improved third-grade reading scores across the school district.

Before the program, third-grade students in the district averaged 72.6 points on standardized reading tests with a standard deviation of 4.8 points.

After implementing the program for one semester, you collected scores from 12 randomly selected classrooms:
74, 76, 73, 75, 78, 77, 74, 79, 75, 76, 77, 75

As a journalist, you need to determine: **Is there statistical evidence that reading scores have actually improved?**

## Task 1: Organize your data and initial assessment

Before you can run this codeblock, you will need to fill in a value where it says REPLACE_ME. That value can be found in the introduction.

```{r}
prior_mean <- 72.6 
prior_sd <- 4.8    


score_data <- tibble(
  classroom = paste("Classroom", 1:12),
  reading_score = new_scores
)

score_data
```

### Reflection Question 1:  (2 points)
Based on just looking at the score_data dataframe, have test scores improved? How can you tell?

Yes they have improved. You can tell because every classroom’s reading score is higher than the previous district average of 72.6. The new scores range from 73 to 79, with most classrooms scoring around 75–77, which is several points above the earlier average.

## Task 2: Calculate key statistics  (2 points)

Like Task 1, you will need to replace values where it says REPLACE_ME before running any code.


```{r}

new_stats <- score_data |> 
  summarise(
    mean = mean(reading_score),
    sd = sd(reading_score),
    n = n()
  )

new_stats
```

### Reflection Question 2: (3 points)
Looking at the mean and standard deviation of the new scores compared to the previous statistics, what initial observations can you make? Provide examples from the data in your answer. What questions might these statistics raise for your reporting?

The new average score is 75.75, which is about 3 points higher than the old average of 72.6. That means students are doing better overall after the new reading program. The standard deviation went down from 4.8 to 1.76. This means most classrooms scored around the same level, and the improvement happened kind of evenly.

Some questions are:
- Did students take the same kind of test as before?
- Could other things, like smaller classes or extra reading help, have affected the results?

## Task 3: Create a column chart  (1 points)

As before, replace any values marked REPLACE_ME based the instructions.


```{r}
my_fill_color <- "darkgreen" 

score_data |> 
ggplot(aes(x = classroom, y = reading_score)) +
  geom_col(fill = my_fill_color, alpha = 0.8) +
  geom_hline(yintercept = prior_mean, color = "darkred", size = 1, linetype = "dashed") +
  annotate("text", x = 2, y = prior_mean - 1, 
           label = "Prev. average (72.6)",  hjust = -0.5, vjust = 2, fontface = "bold", color = "darkred") +
  labs(
    title = "Reading scores after new program implementation",
    subtitle = "Horizontal line shows prev. district average of 72.6 points",
    x = NULL,
    y = "Reading test score",
    caption = "Source - district assessment data"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

### Reflection Question 3: (3 points)
Examine the chart you created, and suggest a better title based on the results of the data, not a description.

New reading program helps 3rd graders score higher across classrooms. 

## Task 4: Perform a hypothesis test (2 points)

This is where we formally test the superintendent's claim that reading scores have improved. Fill in the REPLACE_ME values as needed, beginning with your hypotheses.

**Hypotheses:**
Null: The new reading program REPLACE_ME
Alternative: The new reading program REPLACE_ME

```{r}
alpha_level <- 0.05  

t_test_result <- t.test(
  score_data$reading_score,
  mu = prior_mean,
  alternative = "greater"
)

t_test_result
```

### Reflection Question 4:  (3 points)
What does the p-value tell you, and what doesn't it tell you? How would you explain these results to a non-technical audience while maintaining accuracy?

The p value shows that there’s a small chance the higher scores happened just by luck if the average was still 72.6. Because it’s much smaller than 0.05, we can say the new reading program likely led to a real improvement in scores.

But, the p value doesn’t tell us how big the improvement is or what caused it. It just shows that the increase is probably not due to chance.

## Task 5: Interpreting the results for your news story (2 points)

Let's gather all of the important stats we'll need in one place, so we can look at the prior average, the new scores and the results of the t.test, including the confidence interval. Replace any values where it says REPLACE_ME.


```{r}
p_value <- t_test_result$p.value

ci <- t.test(score_data$reading_score)$conf.int

story_stats <- tibble(
  `Previous average` = prior_mean,
  `New average` = mean(score_data$reading_score),
  `Improvement` = mean(new_scores) - prior_mean,
  `Percent change` = round(((mean(new_scores) - prior_mean) / prior_mean) * 100, 1),
  `p-value` = p_value,
  `Lower bound` = ci[1],
  `Upper bound` = ci[2],
  `Confidence level` = "95%"
)

story_stats
```

## Conclusion

### Reflection Question 5: (4 points)
Based on these statistics, what would be your headline and lead paragraph for this story? Is there evidence to support the superintendent's claim?

Headline: New reading program leads to significant score increase in multiple classrooms

Lede: After the district started a new reading program, third-graders test scores went up. The average score rose from 72.6 to 75.8, which is a clear improvement. A statistical test shows the increase is real and not just random, supporting the superintendent’s claim that the program is working.

### Reflection Question 6:  (1 point)
What metrics or outcomes beyond test scores might be important to track for assessing reading performance?

Besides test scores, it’s important to look at other ways students are improving. This can include: how well they understand what they read, how confident they are reading aloud, and whether they enjoy reading more. Teachers and parents can also give feedback on students’ reading habits to see if the program is actually helping.

