
---
title: "Activity: Linear Regression "
date: 11-17-2025
Dariana Guzman
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

## Identifying Schools with Biggest Differences in Math Test Scores

### Introduction

In this analysis, we will create a scatterplot to visualize the relationship between the FARMS (Free and Reduced-price Meal System) percentage and the percentage of students who met or exceeded math expectations. We'll then use linear regression to identify the schools that show the biggest differences between their actual and expected math test scores.

This analysis can be useful for journalists to understand which schools are performing better or worse than expected, based on the socioeconomic status of their student population. By identifying these schools, targeted interventions and support can be provided to help improve student outcomes - and to provide some context on comments by policymakers.

### Loading the Data

First, let's load the data into a data frame using the read_csv() function from the tidyverse library.

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_pa.csv")
```
### Task: Create a Scatterplot and Describe the Results (2 points)

Next, let's create a scatterplot to visualize the relationship between the FARMS  (Free and Reduced-price Meal System) percentage and the percentage of students who met or exceeded math expectations. We'll use ggplot2 to create the plot and add a linear regression line.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations", title = "Scatterplot of Math Expectations vs FARMS Percentage")
```

The scatterplot shows the overall trend of the relationship between the FARMS percentage and the math test scores. The red line represents the linear regression line, which helps us understand the average relationship between the two variables. Based on what you see here, what's the relationship between FARMS percentage and math test scores?

**WRITE YOUR ANSWER HERE**

The scatterplot shows that there isn’t a strong relationship between the FARMS percentage and math test scores. 

The points are spread out, which means that schools with both high and low FARMS percentages can have either high or low math performance. 

The regression line slopes upward slightly, which can show a very small positive trend, but it’s so weak that the FARMS percentage doesn’t really predict how well students do on the math test. 

Overall -- the graph shows that math scores vary a lot, no matter what the FARMS percentage is.

### Task: Linear Regression Model (3 points)

Next, we'll perform a linear regression analysis using the lm() function. We'll set the percentage of students who met or exceeded math expectations as the dependent variable, and the FARMS percentage as the independent variable. In other words, if we know the FARMS percentage for a school, how reliably can we predict its math score?

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)

summary(model)
```

There are three things we need to examine here: the p-value, the Adjusted R-squared value and the coefficients. You know what a p-value is and how to evaluate it. The Adjusted R-squared is a measure of how much of the difference between schools' math scores can be explained by their FARMS percentage. You don't need to multiply that figure by anything: that's a percentage. The closer the adjusted R-squared is to 100, the more of the outcome the independent variable explains. If both the p-value and adjusted R-squared values are significant, we can look to the coefficients.

In the middle, you can see the (Intercept) is 59.57897 and the FARMS coefficient is 0.12349. Remember high school algebra? Remember learning the equation of a line? Remember swearing that learning y=mx+b is stupid because you’ll never need it again? Surprise. It’s useful again. In this case, we could try to predict a school's score math score by using y=mx+b. In this case, y is the math score, m is 0.12349 and b is 59.57897. So we would multiply a school's FARMS percentage by 0.12349 and then add 59.57897 to it. The result would tell you what the math score would be, according to our model.

Based on what you've learned so far, is this model worth using? What does it tell us about the nature of the relationship between FARMS percentage and math scores?

**WRITE ANSWER HERE**

This model is not worth using because it does not do a good job explaining or predicting math scores. 

The p-value is not significant, which means the relationship between FARMS percentage and math scores could easily be due to chance, the student. 

The adjusted R-squared is extremely low, only about 2-3% so FARMS percentage explains almost none of the differences in math scores between schools. Even though we can plug values into the y = mx + b equation, the model isn’t reliable because the connection between the two variables is so weak. 

So, this tells us that FARMS percentage does not strongly affect or predict math performance in this dataset, and math scores vary widely no matter what the FARMS percentage is.

### Task: Identifying Schools with Biggest Residuals (2 points)

To find the schools with the biggest differences between their actual and expected math test scores, we'll extract the residuals from the linear regression model. The residuals represent the difference between the actual and predicted values. Residuals can tell you several things, but most important is if a linear model the right model for your data. If the residuals appear to be random, then a linear model is appropriate. If they have a pattern, it means something else is going on in your data and a linear model isn’t appropriate.

Residuals can also tell you who is under-performing and over-performing the model. And the more robust the model – the better your r-squared value is – the more meaningful that label of under or over-performing is. 

We'll then use the arrange() function to sort the data frame by the absolute value of the residuals.

```{r}
schools$predicted <- predict(model)

# Calculate the residuals
schools <- schools |> 
  mutate(residual = `% Met or Exceeded Math Expectations` - predicted)

# Sort the data frame by the absolute value of the residuals in descending order
schools <- schools |>
  mutate(abs_residual = abs(residual)) |>
  arrange(desc(abs_residual))

```
The schools with the biggest positive residuals are performing better than expected, while the schools with the biggest negative residuals are performing worse than expected, based on their FARMS percentage. Look at the absolute values of the residuals; what are the largest differences between the predicted result and the actual math scores?

**WRITE YOUR ANSWER HERE**

The largest differences between the predicted math scores and the actual scores come from the schools with the biggest absolute residuals. 

The biggest gap is at Whitemarsh Elementary School -- which has an absolute residual of 34.9 points, which means its actual score is almost 35 points lower than what the model predicted. 

Other schools with very large negative gaps include Gulph Elementary (-31.3), Lower Gwynedd (-30.5), and Rydal West (-29.2). These schools are performing much worse than expected based on their FARMS percentages. 

On the positive side, the school performing the best is Crooked Billet Elementary, with a residual of 31.1, meaning its actual score is about 31 points higher than predicted. Woodland Elementary (+26.6) and Upper Moreland Elementary (+24.8) also score well above what the model suggests.

The biggest differences positive and negative range from about 25 to 35 points, show that some schools are doing far better or far worse than the model would expect.


## Task: Evaluate the model (3 points)

A school board member has said at a meeting that elementary schools with fewer students on free and reduced price meals do better on standardized math tests. Describe what conclusions you'd draw from this model's results and how it would influence a story pitch about the school board member and math scores.

**WRITE YOUR ANSWER HERE**
The results from the model show that the school board member’s claim, that schools with fewer students on free and reduced price meals do better on math tests, is not supported by the data. 

The relationship between FARMS percentage and math scores is not statistically significant, and the adjusted R-squared value shows that FARMS explains almost none of the differences in scores between schools. 

Some high FARMS schools perform far better than predicted, while several low FARMS schools score far below expectations. 

This could mean that poverty levels alone are not a reliable predictor of math performance, and relying on that assumption can oversimplify a much more difficult issue. A story based on these findings would challenge the board member’s claim and highlight how local data shows a far more nuanced picture of school performance.

