---
title: "Oct 1 Activities: M&Ms; Survey; Sampling"
date: 9-30-2025
Name: Dariana Guzman
---

All three class activities are on this .rmd notebook

Activity #1: Activity: M&Ms - Central Limit Theorem

Working in groups, each group will get a handful of candy from the bag, and then count the number of red and black candies. Log those in this Google Sheet 

https://docs.google.com/spreadsheets/d/1KOeoVTq2W7L2T9sWjwh5U_2f59GsTeAylVGD53rNWww/edit?usp=sharing

(give your group a nice name).

Take a look at the values that we all get - what do you notice about the patterns? This is an exercise about the central limit theorem. 

The central limit theorem states that the sampling distribution of the mean of any independent, random variable will be normal or nearly normal, if the sample size is large enough.

Then discuss and try to answer these questions:

--How close is your group's mean to the true proportion of 0.5?
--Why do different groups get different results?
--What happens to individual extreme results when we combine all groups?
--How does this relate to polls and surveys in real life?

Write your answer below. And then you may eat the M&Ms.

**A:**
1. After viewing the sample, the mean to the true proportion was not close at all. There was a majority of colored M&Ms to the point it made it difficult for it to be remotely close to the half way mark.  

2. Sample sizes are different for each group, so there is an inconsistency across the data if viewed by everyone who was given a different sample size. It's just a random sample, so it's not a controlled group. Additionally, it is far too small.

3. If we combine all groups, then there would be a more neutral average which could lead us closer to the true proportion. It would be a bigger group and it would be more controlled because we're taking everything into account. Not just one small sample. It would be more reliable for the entire bag and not just for one specific sample size.

4. This relates to polls and surveys in real life because we aim to have accurate data by having a large and diverse sample size. All of it to prevent any errors or bias and to have a better understanding of the entire population -- not just one set of individuals. 

------------------------------------------------------------------------------

# Activity #2:

Belonging Survey

Working in pairs, take a look back at the main findings from UMD's Belonging Survey:

https://president.umd.edu/internal/belonging


Specifically, focus on the "Sample Description" section and the Appendix. Discuss the differences and between the population and the sample, and what the document says about how the survey attempted to deal with them. Are there any decisions described in the document that raise questions in your minds? Detail the differences you see and your understanding of the methods they employed (it's ok if you don't understand all of them), and add your questions. You should discuss this as a pair but submit individually.

**A:**

Only 12,233 members of the UMD community responded to the survey. 

The quantitative data was analyzed to identify significant differences in key climate indicators across demographic groups using a combination of t-tests and ANOVAs – all depending on variable type. With over 25,000 qualitative responses, they were analyzed using content analysis to extract common themes and thematic patterning across respondents, so that is how they managed to differentiate the population and the sample. It also includes only statistically significant group differences for the demographic variables, while offering important insights into how different groups experienced belonging and community on campus.

I have a few questions regarding the confirmatory factor analysis. They made sure to assure that the scale scores were not designed to be objective measures – but how can be really ensure that this is the case? How can we ensure that they are not being “normed against any population.” Bias is easy to include in any kind of survey or work – keeping it clear of it can be an issue. There is also the fact that not everyone applies to one specific group, how can we also ensure that they are involved? They collapsed ethnic and racial groups together – which can impact the experiences of everyone there. I also want to point out that the sample size is an issue as it doesn’t even cover half of the campus/community population. 

------------------------------------------------------------------------------

# TASK 3: SAMPLING

Based on the class sampling demo (available on Elms), sample data from a population using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

**A**
```{r}
library(tidyverse)

```


### Task 2: Load 1,000 rows of data with the ages of people from 18-80

**A**
```{r}

set.seed(98)
people <- tibble(
  age = sample(18:80, 1000, replace = TRUE)
)

head(people)

```

### Task 3: Sample 100 rows of data from the dataset
**A**
```{r}

sample_100 <- people |> 
  sample_n(100)

```


### Task 4: Calculate the mean age and standard deviation of the sample
**A**
```{r}
sample_100 |> summarize(mean = mean(age), median = median(age), sd = sd(age))

```


### Task 5: Calculate the mean age and standard deviation of the population
**A**
```{r}
people |> summarize(mean = mean(age), median = median(age), sd = sd(age))


```


### Task 6: Compare the sample and population statistics
**A**

The sampl statistics  determine that the mean is 48.74 with a median of 50 and standard deviation of 16.2. The population statistics determine that the mean is 48.8 with a mmedian of 49 and a standard deviation of 17.7. They are very similar to each other.

Compare the means and standard deviations. Are they similar? Go back and re-run the sampling and statistics calculations with a smaller sample size. How do the results change?

**A**:

The means and the standard deviations are quite similar to each other. The results won't change that much when calculating a smaller sample size. The men always fall on or under 50%. 

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
